some title

### 2018-01-04 term paper_2018/paper.tex
correlation in simple lin reg

\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\card}{card}
\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\Corr}{Corr}
\DeclareMathOperator{\E}{E}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\I}{\mathbb{I}} % индикатор события
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

@@

The most crucial part in defining correlation geometrically is definig the dot product as it enables to compute the length of a vecotr:
\[
|\vec{a}| = \sqrt{\langle \vec{a},  \vec{a}\rangle}
\]
and the angle between any two vectors:
\[
\cos(\vec{a}, \vec{b}) = \frac{\langle \vec{a},  \vec{a}\rangle}{|\vec{a}| |\vec{b}|}
\]
Now we define scalar product of two random vectors as covariation between them:
\[
\langle X, Y \rangle = \Cov(X, Y)
\]
The main characteristics of a random vector are its length and direction.
So, we introduce the length
\[
\sqrt{\Cov{X,X}} = \sqrt{\Var(X)} = \sigma_X
\]
and the angle between two random vectors
\[
\cos(X,Y) = \frac{\Cov(X,Y)}{\sqrt{\Var(X)\Var(Y)}} = \Corr(X,Y)
\]
Note that from the definition of the angle it follows that correlation is limited within an interval $[-1,1]$.

@@

Another important geometrical tool is projection. Recall that for any two vectors the scalar product $\langle \vec{a}, \vec{b} \rangle$ can be interpreted as the length of projected $\vec{b}$ multuplied by the length of $\vec{a}$.
The projection itself is $cos(\vec{a}, \vec{b}) \vec{b}$.
Same holds for random vectors. The projection of a random vector $Y$ onto $\{cX| c \in \mathbb{R}\}$ is $\Corr(X,Y) \cdot Y$.

@@

\begin{figure}[ht!]
\begin{center}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/averages_1.pdf}
%\caption{Regression of $y$ on $Lin(x,z)$} %% подпись к рисунку
\label{fig:averages_lin} }
\hspace{4ex}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/averages_2.pdf}
%\caption{$Lin(x, z)$}
\label{fig:averages_bars}}
\caption{\subref{fig:averages_lin}: Regression of $y$ on $Lin(\mathbb{1},x)$; \subref{fig:averages_bars}: Regression of $y$ and $\hat y$ on $\mathbb{1}$.}
\end{center}
\end{figure}

>>>>

\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\sVar}{sVar}
\DeclareMathOperator{\card}{card}
\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\Corr}{Corr}
\DeclareMathOperator{\sCov}{sCov}
\DeclareMathOperator{\sCorr}{sCorr}
\DeclareMathOperator{\E}{E}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\I}{\mathbb{I}} % индикатор события
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

@@

The most crucial part in defining correlation geometrically is definig the dot product as it enables to compute the length of a vecotr:
\[
|\vec{a}| = \sqrt{\langle \vec{a},  \vec{a}\rangle}
\]
and the angle between any two vectors:
\[
\cos(\vec{a}, \vec{b}) = \frac{\langle \vec{a},  \vec{a}\rangle}{|\vec{a}| |\vec{b}|}
\]
Now we define scalar product of two random vectors as covariation between them:
\[
\langle X, Y \rangle = \Cov(X, Y)
\]
The main characteristics of a random vector are its length and direction.
So, we introduce the length
\[
\sqrt{\Cov{X,X}} = \sqrt{\Var(X)} = \sigma_X
\]
and the angle between two random vectors
\[
\cos(X,Y) = \frac{\Cov(X,Y)}{\sqrt{\Var(X)\Var(Y)}} = \Corr(X,Y)
\]
Note that from the definition of the angle it follows that correlation can range from $-1$ to $1$.

@@

Another important geometrical tool is projection. Recall that for any two vectors the scalar product $\langle \vec{a}, \vec{b} \rangle$ can be interpreted as the length of projected $\vec{b}$ multuplied by the length of $\vec{a}$.
The projection itself is $cos(\vec{a}, \vec{b}) \vec{b}$.
Same holds for random vectors. The projection of a random vector $Y$ onto $\{cX| c \in \mathbb{R}\}$ is $\hat Y = \Corr(X,Y) \cdot Y$.

\begin{figure}[ht!]
\begin{center}
\includegraphics{images/corr_proj.pdf}
\label{fig:corr_proj}
\caption{Random vector $Y$ projected onto random vector $X$.}
\end{center}
\end{figure}

Looking at Figure~\ref{fig:corr_proj}, we can interpret the square of correlation coefficient.
Using the fact that $cos^2 \varphi$ is the squared ratio of the leg adjacent to $\varphi$ to hypotenuse, we can write
\[
\Corr^2(X,Y) = \frac{\Var(\hat Y)}{\Var(Y)}
\]
as the variance of a random vecor is associated with the square of its length.
Thus, the correlation coefficient squared shows the fraction of variance in $Y$ which can be explained
with the most similar random variable proportional to $X$.

\subsection{Sample correlation coefficient in simple linear regression}

\marginnote{
Assuming the underlying relationship between $x$ and $y$ to be
\[
y_i = \beta_1 + \beta_2 x_i + \varepsilon_i \quad i=1,\ldots,n
\]
where $\varepsilon_i$ is an error term the following holds
\begin{align*}
\sCorr(y, \hat y) &= \frac{\sCov(y)\sCov(\hat y)}{\sqrt{\sVar(y)\sVar(\hat y)}} \\
&= \frac{\sCov(y)\sCov(\hat \beta_1 + \hat \beta_2 x)}{\sqrt{\sVar(y)\sVar(\hat \beta_1 + \hat \beta_2 x)}} \\
&= \frac{\sCov(y)\sCov(\hat \beta_2 x)}{\sqrt{\sVar(y)\sVar(\hat \beta_2 x)}} \\
&= \frac{\hat \beta_2 \sCov(y)\sCov(x)}{|\hat \beta_2| \sqrt{\sVar(y)\sVar(x)}} \\
&= sign(\hat \beta_2) \frac{\sCov(y)\sCov(x)}{\sqrt{\sVar(y)\sVar(x)}}
\end{align*}
}

\begin{theorem}
A linear regression model with one explanatory variable and constant term has the property
\[
\sCorr(y, \hat y) = sign(\hat \beta_2) \sCorr(y, x)
\]
\end{theorem}

\begin{proof}
Firstly, we consider the case when $\hat \beta_2 > 0$ so the main picuture is of the form depicted in Figure~\ref{fig:corr_pos_base}.
It has been shown earlier that the correlation coefficient squared represents the angle betweem two random vectors.

\begin{figure}[ht!]
\begin{center}
\includegraphics{images/corr_eq_positive.pdf}
\label{fig:corr_pos_base}
\caption{Vectors $x$, $y$ and $\mathbf{1}$.}
\end{center}
\end{figure}

However, it seems to be difficult to compare the angles in the three dimensional space.
That is why we start with projecting both $x$ and $y$ onto the space perpendicular to the vector of all ones $\mathbf{1}$ as shown in Figure~\ref{fig:corr_pos_xyc}.
We denote this space as $Lin^{\perp}(\mathbf{1})$. The resulting vectors are $x - \bar x \cdot \mathbf{1}$  and $y - \bar y \cdot \mathbf{1}$ respectively
since projection of any vector $\vec{a}$ on the line given by a vector of all ones yields the vector of averages $\vec{\bar a}$.

In order to get the angle between $y$ and $\hat y$ we should start with regressing $y$ on $Lin(x, \mathbf{1})$.
Then the only thing thing left is to project $\hat y$ onto $Lin^{\perp}\mathbf{1}$ since the $y$ vector has already been projected.
The result of this step is shown in Figure~\ref{fig:corr_pos_yhatc}.

\begin{figure}[ht!]
\begin{center}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/corr_eq_positive_xy_centered.pdf}
\label{fig:corr_pos_xyc}}
\hspace{4ex}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/corr_eq_positive_yhat.pdf}
\label{fig:corr_pos_yhatc}}
\caption{\subref{fig:corr_pos_xyc}: `Centred' $x$ and $y$, i.e., projected onto $Lin^{\perp}(\mathbf{1})$; \subref{fig:corr_pos_yhatc}: `Centred' $\hat y$, i.e., projected onto $Lin^{\perp}(\mathbf{1})$.}
\end{center}
\end{figure}

todo: whole picture, $\beta_2<0$

@@

\begin{figure}[ht!]
\begin{center}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/averages_1.pdf}
\label{fig:averages_lin} }
\hspace{4ex}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/averages_2.pdf}
\label{fig:averages_bars}}
\caption{\subref{fig:averages_lin}: Regression of $y$ on $Lin(\mathbb{1},x)$; \subref{fig:averages_bars}: Regression of $y$ and $\hat y$ on $\mathbb{1}$.}
\end{center}
\end{figure}
